---
title: rawrr - Functional Test using raw Files (autoQC01) from Different LC-MS
  Systems
author: "Tobias Kockmann and Christian Panse"
output: pdf_document
abstract: |
  This Rmarkdown script processes raw files recorded across different Thermo
  Scientific MS at the [FGCZ proteomics unit](https://fgcz.ch/omics_areas/prot.html)
  using the latest `R` development and [rawrr](https://github.com/fgcz/rawrr)
  package versions.
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(mc.cores = 8)
```
\newpage

# Objectives

The main objectives are:

* Demonstrate code robustness with respect to different instrument models.
* Showcase dynamic application (daily changing input data).
* Demonstrate multicore computation by applying the `parallel` package.

The automatically updated output of this R script can be viewer at https://fgcz-ms.uzh.ch/~cpanse/rawR/test/functional_test.html

Of note, The analyzed sample (termed autoQC01) consists of the iRT peptide mix (Biognosys) in a tryptic BSA digest (NEB) and was separated applying a 20 min linear gradient on C18 RP material at a constant flow rate of 300 nl/min.


# Download raw files

A static snapshot of raw files was deposited as MassIVE dataset [MSV000086542](https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?accession=MSV000086542).
For local computation please download the raw files using ftp://massive.ucsd.edu/MSV000086542/.

```
wget ftp://massive.ucsd.edu/MSV000086542/raw/*.raw
```

# Load R packages

```{r library}
library(rawrr)
library(parallel)

# only used for computing PSMs
library(protViz)
```

# List raw files

```{r list}
(input <- list.files(patter=".*raw$"))
```

# Check/demonstrate `rawrr` reader functions

Below we check the functionality of all reader functions of the `rawrr` package.

|Function Name|Description|Return value|
|:------------|-------------------------------|:------------------|
|`readFileHeader`|Reads meta information from a rawfile header.|`list`|
|`readIndex`|Reads scan index from a rawfile.|`data.frame`|
|`readSpectrum`|Reads spectral data from a rawfile|`rawRspectrum(Set)`|
|`readChromatogram`|Extracts chromatograms (TIC, BPC, or XIC) from a rawfile|`rawRchromatogram(Set)`|

## `readFileHeader`

Read a list of meta data for each raw file.

```{r fileHeader}
H <- mclapply(FUN=rawrr::readFileHeader, input)
```

```{r defineInfo, echo=FALSE}
.getInstrumentInformation <- function(x){
  data.frame(
             model=x$`Instrument model`,
             serialNumber=x$`Serial number`,
             #method=x$`Instrument method`,
             softwareVersion=x$`Software version`,
             nMS2=x$`Number of ms2 scans`)
}
```

The table below lists some instrument information:

```{r knitr, results='asis'}
knitr::kable(instrumentInformation<-do.call('rbind', lapply(H, .getInstrumentInformation)))
```

## `readIndex` and `readSpectrum`


Defines iRT peptide m/z values and scores.

```{r defineIrtMass}
iRT.mZ <- c(487.2571, 547.2984, 622.8539, 636.8695, 644.8230, 669.8384, 683.8282,
            683.8541, 699.3388, 726.8361, 776.9301)

iRT.score <- c(-24.92, 19.79, 70.52, 87.23, 0, 28.71, 12.39, 33.38, 42.26, 54.62, 100)

names(iRT.mZ) <- c("LGGNEQVTR", "YILAGVENSK", "GTFIIDPGGVIR", "GTFIIDPAAVIR",
                 "GAGSSEPVTGLDAK", "TPVISGGPYEYR", "VEATFGVDESNAK",
		 "TPVITGAPYEYR", "DGLDAASYYAPVR", "ADVTPADFSEWSK",
		 "LFLQFGAQGSPFLK")
```



```{r defineComputeBestPeptideSpectrumMatch, echo=FALSE}
.computeBestPeptideSpectrumMatch <- function(rawfile,
   pepSeq = c("LGGNEQVTR", "GAGSSEPVTGLDAK", "VEATFGVDESNAK",
              "TPVISGGPYEYR", "YILAGVENSK", "TPVITGAPYEYR",
              "DGLDAASYYAPVR", "ADVTPADFSEWSK", "GTFIIDPGGVIR",
              "GTFIIDPAAVIR", "LFLQFGAQGSPFLK"),
   peptideMassTolerance = 0.003,
   fragmentIonMassTolerance = 0.01) {
  
  mass2Hplus <- (parentIonMass(pepSeq) + 1.008) / 2
  
  S <- rawrr::readIndex(rawfile)
  S <- S[-which(S$MSOrder != "Ms2"), ]
  if (nrow(S) == 0) return (NULL)
  
  idx <- lapply(mass2Hplus, function(m){
    which(abs(as.double(S$precursorMass) - m) < peptideMassTolerance)
  })
  
  # just to be generic correct
  scanNumbers <- lapply(idx, function(x){S$scan[x]})
  
  yIons <- function (b, y) return(y)
  
  bestMatchingMS2Scan <- mclapply(1:length(pepSeq), function(i){
    PL <- rawrr::readSpectrum(rawfile, scan = scanNumbers[[i]])
    
    pp <- lapply(PL, function(x){protViz::psm(pepSeq[i], x, FUN = yIons, plot = FALSE)})
    
    intensity.max <- max(sapply(pp, function(x) sum(x$intensity)))
    
    ## derive scores for each match
    score <- sapply(1:length(pp),
                    function(j){                                                                                                   
                      ## TODO: implement some more sophisticated score from the literature
                      ## intensityOfMatches <- sum(PL[[j]]$intensity[abs(pp[[j]]$mZ.Da.error) < 0.1])
                      ## nomalizedIntensityOfMatches <- median(PL[[j]]$intensity[abs(pp[[j]]$mZ.Da.error) < 0.01]) / intensity.max
                      
                      # hard constraint
                      nIonHits <- sum(abs(pp[[j]]$mZ.Da.error) < fragmentIonMassTolerance)
                      nIonHits
                    }) 
    ## determine best scoring spectra
    bestFirstMatch <- which(max(score, na.rm = TRUE) == score)[1]
    
    ## return the winner
    data.frame(scan=scanNumbers[[i]][bestFirstMatch], peptide=pepSeq[i], rawfile=rawfile)
  })
  
  do.call('rbind', bestMatchingMS2Scan)
}
```
The R function `.computeBestPeptideSpectrumMatch` defines some basic heuristic to determine peptide to spectrum matches for the iRT peptides.

Perform the peptide spectrum match number crunching for the `r names(iRT.mZ)[1]` peptide on all raw files:

```{r computeBestMatches}
start_time <- Sys.time()
bestMatchingMS2Scan <- mclapply(FUN = .computeBestPeptideSpectrumMatch,
                                input, pepSeq=names(iRT.mZ)[1])
end_time <- Sys.time()
end_time - start_time
```
## `plot` spectrum

```{r definePlotBestMatches, echo=FALSE}
.plotBestMatches <- function(x, instrumentModel){
  if(is.null(x)) return (NULL)
  S <- readSpectrum(x$rawfile[1], x$scan)
  
  lapply(1, function(i){
    plot(S[[i]], main=x$peptide[i], sub=instrumentModel)
    Y <- x$peptide[i]
    abline(v=fi<-fragmentIon(Y)[[1]]$y[1:nchar(Y)-1],
           col="#AAAAAA66",lwd=5)
    
    axis(3, fi, paste0('y', 1:length(fi)))
  })}
```

```{r plotBestMatches, fig.show="hold", out.width="33%"}
start_time <- Sys.time()
op <- par(mar = c(6, 6, 6, 1), mfrow=c(1, 1))
rv <- mapply(FUN=.plotBestMatches, bestMatchingMS2Scan,
             instrumentInformation$model)
end_time <- Sys.time()
end_time - start_time
```

## `readChromatogram`

Derives chromatograms for all iRT peptides applying the mass filter option and an ion tolerance of 10ppm:

```{r readChromatogram, error=TRUE}
start_time <- Sys.time()
C <- mclapply(X=input, FUN=rawrr::readChromatogram,
              mass = iRT.mZ, tol = 10, type = "xic", filter = "ms")
end_time <- Sys.time()
end_time - start_time
```
## `plot` chromatogram

Visualize the chromatograms and apply a linear regression:

```{r definePlotChromatogramAndFit, echo=FALSE}
.plotChromatogramAndFit <- function(x, i){
    par(mfrow=c(2,1))
    
    plot(x); legend("topright", legend=i, title='Instrument Model', bty = "n", cex=0.75)
    
    rt <- sapply(x, function(x) x$times[which.max(x$intensities)[1]])
    if (length(rt) == length(iRT.score)){
    fit <- lm(rt ~ iRT.score)
    plot(rt ~ iRT.score, ylab = 'Retention time [min]',
         xlab = "iRT score", pch=16, frame.plot = FALSE)
    abline(fit, col = 'grey')
    abline(v = 0, col = "grey", lty = 2)
    legend("topleft",
     legend = paste("Regression line: ", "rt =",
                    format(coef(fit)[1], digits = 4), " + ",
                    format(coef(fit)[2], digits = 2), "score", "\nR2: ",
                    format(summary(fit)$r.squared, digits = 2)),
           bty = "n", cex = 0.75)
    text(iRT.score, rt,  iRT.mZ,pos=1,cex=0.5)
    }
}
```

```{r plotChromatogramAndFit, fig.retina=3,fig.height=9,  fig.show="hold", out.width="33%"}
rv <- mapply(FUN=.plotChromatogramAndFit,
             x=C, i=instrumentInformation$model)
```

# Listings

## `.getInstrumentInformation`
```{r}
.getInstrumentInformation
```

## `.computeBestPeptideSpectrumMatch`
```{r}
.computeBestPeptideSpectrumMatch
```

## `.plotBestMatches`
```{r}
.plotBestMatches
```

## `.plotChromatogramAndFit`

```{r}
.plotChromatogramAndFit
```

# Session information


```{r monoInfo, message=TRUE, echo=TRUE}
rawrr:::.monoInfo()
```

```{r sessionInfo}
sessionInfo()
```


